---
title: "Midterm Result" 
author: "Yiying Wu (yw3996)"
output:
  pdf_document:
header-includes:
- \usepackage{fancyhdr}
- \usepackage{lipsum}
- \pagestyle{fancy}
- \fancyhead[L]{Data Science 2 Midterm}
- \fancyhead[R]{Yiying Wu (yw3996)}
- \fancypagestyle{plain}{\pagestyle{fancy}}
--- 

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
```


```{r}
## R packages

library(tidyverse)
library(caret)
library(tidymodels)
library(gtsummary)
library(corrplot)
```

## Input dataset
```{r}
# Load the .RData file
load("./recovery.RData")
```

```{r}
## recode categorical data
dat <-dat %>% 
  mutate(
    ## gender
    male=gender,
    ## race
    race=relevel(as.factor(race), ref ="1"),
    ## smoking
    smoking=relevel(as.factor(smoking), ref ="0"),
    ## study
    study=relevel(as.factor(study), ref ="A")
  )
```


## Exploratory analysis and data visualization

In this section, use appropriate visualization techniques to explore the dataset and identify any patterns or relationships in the data.

**Summary statistics**
```{r}
summ_dat<-dat%>%
  select(-id,-male)%>%
  mutate(
    gender = factor(case_when(
      gender == "1" ~ "female",  
      gender == "0" ~ "male"),
      levels = c("male", "female")
    ),
    race= factor(case_when(
      race == "1" ~ "White",  
      race == "2" ~ "Asian",
      race == "3" ~ "Black",
      race == "4" ~ "Hispanic"),
      levels = c("White", "Asian","Black","Hispanic")
    ),
    smoking=factor(case_when(
      smoking == "0" ~ "Never smoked",  
      smoking == "1" ~ "Former smoker",
      smoking == "2" ~ "Current smoker"),
      levels = c("Never smoked", "Former smoker","Current smoker")
    ),
    hypertension=factor(case_when(
      hypertension == "0" ~ "No",  
      hypertension == "1" ~ "Yes"),
      levels = c("No", "Yes")
    ),
    diabetes=factor(case_when(
      diabetes == "0" ~ "No",  
      diabetes == "1" ~ "Yes"),
      levels = c("No", "Yes")
    ),
    vaccine=factor(case_when(
      vaccine == "0" ~ "Not vaccinated",  
      vaccine == "1" ~ "Vaccinated"),
      levels = c("Not vaccinated", "Vaccinated")
    ),
    severity=factor(case_when(
      severity == "0" ~ "Not severe",  
      severity == "1" ~ "Severe"),
      levels = c("Not severe", "Severe")
    )
    )

summ_dat %>% 
  tbl_summary() %>% 
  bold_labels()%>%
  as_gt(include = everything()) %>%
  gt::tab_header("Table 1: Summary of Dataset")
```

**Visualizations for the numerical variables**

```{r}
ggplot(summ_dat, aes_string(x = "age")) + geom_histogram(bins = 30) + ggtitle("Histogram of age")

ggplot(summ_dat, aes_string(x = "height")) + geom_histogram(bins = 30) + ggtitle("Histogram of height")

ggplot(summ_dat, aes_string(x = "weight")) + geom_histogram(bins = 30) + ggtitle("Histogram of weight")

ggplot(summ_dat, aes_string(x = "bmi")) + geom_histogram(bins = 30) + ggtitle("Histogram of bmi")

ggplot(summ_dat, aes_string(x = "SBP")) + geom_histogram(bins = 30) + ggtitle("Histogram of SBP")

ggplot(summ_dat, aes_string(x = "LDL")) + geom_histogram(bins = 30) + ggtitle("Histogram of LDL")

ggplot(summ_dat, aes_string(x = "recovery_time")) + geom_histogram(bins = 30) + ggtitle("Histogram of recovery_time")
```
**correlation plot**

```{r}
numerical_data <- summ_dat[, sapply(summ_dat, is.numeric)]
correlation_matrix <- cor(numerical_data)
corrplot(correlation_matrix, method = "circle")
```


## Model training

In this section, describe the models you used to predict the time to recovery from COVID-19. Briefly state the assumptions made by using the models. Provide a detailed description of the model training procedure and how you obtained the final model.

### MARS
```{r}
set.seed(666)
data_split <- initial_split(dat, prop = 0.8)
```


### GAM

### lasso

### Elastic net 

ctrl1 <- trainControl(method = "cv", number = 10) tuneGrid = expand.grid(alpha = seq(0, 1, length = 21), lambda = exp(seq(-25, 5, length = 100)))