---
title: "midterm ds2"
author: "Tingyi Li"
date: "2024-03-22"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(ISLR)
library(glmnet)
library(caret)
library(tidymodels)
library(pls)
library(earth)
library(bayesQR)
library(ggplot2)
library(pROC)
library(plotmo)
library(corrplot)
library(patchwork)
```

```{r}
load("./recovery.RData")
data <- subset(dat, select = -id)
data$study <- ifelse(data$study == "A", 0, 1)

data <- data |>  
  mutate(
    gender = as.factor(gender),
    race = as.factor(race),
    smoking = as.factor(smoking),
    hypertension = as.factor(hypertension),
    diabetes = as.factor(diabetes),
    vaccine = as.factor(vaccine),
    severity = as.factor(severity),
    study = as.factor(study)
    )
```


# Data Partition
Divide data into training data (80%) and testing data (20%)
```{r}
set.seed(666)
data_split <- initial_split(data, prop = 0.8)

# training data
training_data <- training(data_split)
training_data <- na.omit(training_data)


# testing data
testing_data <- testing(data_split)
testing_data <- na.omit(testing_data)

# matrix of predictors
x <- model.matrix(recovery_time ~ ., training_data) [ ,-1]
y <- training_data$recovery_time

x2 <- model.matrix(recovery_time ~ ., testing_data) [ ,-1]
y2 <- testing_data$recovery_time

# 10-fold cross validation
ctrl1 <- trainControl(method = "repeatedcv", repeats = 10)
```

# EDA
```{r}
skimr::skim(data)
```


## Feature plot for continuous variables 
```{r}
theme1 <- trellis.par.get()
theme1$plot.symbol$col <- rgb(.2, .4, .2, .5)
theme1$plot.symbol$pch <- 16
theme1$plot.line$col <- rgb(.8, .1, .1, 1)
theme1$plot.line$lwd <- 2
theme1$strip.background$col <- rgb(.0, .2, .6, .2)
trellis.par.set(theme1)

trainData <- training_data|>
  dplyr::select('age', 'height', 'weight', 'bmi', 'SBP', 'LDL', 'recovery_time')
featurePlot(x = trainData[ , 1:6],
            y = trainData[ , 7],
            plot = "scatter",
            span = .5,
            labels = c("Predictors","Recovery Time"),
            main = "Figure 1:Relationship between Continuous Predictors and Recovery Time",
            type = c("p", "smooth"),
            layout = c(3, 2))
```

# Correlation plot
```{r}
corr_data = training_data |>
  dplyr::select('age', 'height', 'weight', 'bmi', 'SBP', 'LDL')

corrplot(cor(corr_data), method = "circle", type = "full")
```


## Boxplot for categorical variables
```{r}
gender_plot <- data |>
  ggplot(aes(x = gender, y = recovery_time, fill = gender))+
  geom_violin() +
  scale_fill_discrete(labels = c("female", "male")) +
  labs(
       x = "Gender",
       y = "Recovery_time") +
  theme_minimal()
gender_plot

race_plot <- data |>
  ggplot(aes(x = race, y = recovery_time, fill = race)) +
  geom_violin() +
  scale_fill_discrete(labels = c("White", "Asian", "Black", "Hispanic")) +
  labs(
       x = "race",
       y = "Recovery_time") +
  theme_minimal()
race_plot

smoking_plot <- data |>
  ggplot(aes(x = smoking, y = recovery_time, fill = smoking)) +
  geom_violin() +
  scale_fill_discrete(labels = c("Never smoked", "Former smoker", "Current smoker")) +
  labs(
       x = "smoking",
       y = "Recovery_time") +
  theme_minimal()
smoking_plot 

hypertension_plot <- data |>
  ggplot(aes(x = hypertension, y = recovery_time, fill = hypertension)) +
  geom_violin() +
  scale_fill_discrete(labels = c("No", "Yes")) +
  labs(
       x = "hypertension",
       y = "Recovery_time") +
  theme_minimal()
hypertension_plot

diabetes_plot <- data |>
  ggplot(aes(x = diabetes, y = recovery_time, fill = diabetes)) +
  geom_violin() +
  scale_fill_discrete(labels = c("No", "Yes")) +
  labs(
       x = "diabetes",
       y = "Recovery_time") +
  theme_minimal()
diabetes_plot

vaccine_plot <- data |>
   ggplot(aes(x = vaccine, y = recovery_time, fill = vaccine)) +
   geom_violin() +
   scale_fill_discrete(labels = c("Not vaccinated", "Vaccinated")) +
   labs(
        x = "vaccine",
        y = "Recovery_time") +
   theme_minimal()
vaccine_plot

severity_plot <- data |>
  ggplot(aes(x = severity, y = recovery_time, fill = severity)) +
  geom_violin() +
  scale_fill_discrete(labels = c("Not severe", "Severe")) +
  labs(
       x = "severity",
       y = "Recovery_time") +
  theme_minimal()
severity_plot

study_plot <- data |>
  ggplot(aes(x = study, y = recovery_time, fill = study)) +
  geom_violin() +
  scale_fill_discrete(labels = c("A", "B")) +
  labs(
       x = "study",
       y = "Recovery_time") +
  theme_minimal()
study_plot

gender_plot + race_plot + smoking_plot + hypertension_plot + diabetes_plot + vaccine_plot + severity_plot + study_plot + plot_layout(ncol = 3)
```




# Linear models

## Fit a lasso model
```{r}
set.seed(666)

lasso.fit <- train(recovery_time ~ .,
                   data = training_data,
                   method = "glmnet",
                   tuneGrid = expand.grid(alpha = 1, 
                                          lambda = exp(seq(-25, 5, length = 100))),
                   trControl = ctrl1)
plot(lasso.fit, xTrans = log)

## Select the best tuning parameter
lasso.fit$bestTune

# coefficients in the final model
coef(lasso.fit$finalModel, lasso.fit$bestTune$lambda)
```
The selected tuning parameter is 0.00912288.

Test errors
```{r}
set.seed(666)
lasso.pred <- predict(lasso.fit, newdata = testing_data)
mean((lasso.pred - testing_data[, "recovery_time"])^2)
```

The test error is 298.3016.


## Fit an elastic net model
```{r}
set.seed(666)
enet.fit <- train(recovery_time ~ .,
                  data = training_data,
                  method = "glmnet",
                  tuneGrid = expand.grid(alpha = seq(0, 1, length = 21), 
                                         lambda = exp(seq(-25, 5, length = 100))),

                  trControl = ctrl1)
## Select the best tuning parameter
enet.fit$bestTune

#Plot
myCol <- rainbow(25)
myPar <- list(superpose.symbol = list(col = myCol),
              superpose.line = list(col = myCol))

plot(enet.fit, par.settings = myPar)

# coefficients in the final model
coef(enet.fit$finalModel, enet.fit$bestTune$lambda)

# test error
enet.pred <- predict(enet.fit, newdata = testing_data)
mean((enet.pred - testing_data[, "recovery_time"])^2)
```
The selected tuning parameter is 0.00367552, and the test error is 297.1642.


## Fit a PCR model
```{r}
set.seed(666)
pcr.fit <- train(x, y,
                 method = "pcr",
                 tuneGrid = data.frame(ncomp = 1:15),
                 trControl = ctrl1,
                 scale = TRUE)
predy2.pls2 <- predict(pls.fit, newdata = x2)
mean((y2 - predy2.pls2)^2)

ggplot(pls.fit, highlight = TRUE)
summary(pls.fit)
```
There are 15 components in the model and the test error is 323.779.

# Nonlinear models
## MARS
```{r}
set.seed(666)
mars.grid <- expand.grid(degree = 1:4,
                         nprune = 2:38)
mars.fit <- train(x, y,
                  method = "earth",
                  tuneGrid = mars.grid,
                  trControl = ctrl1)
ggplot(mars.fit)
```
```{r}
mars.fit$bestTune
```
```{r}
coef(mars.fit$finalModel)
# Test error
mars.pred <- predict(mars.fit, newdata = x2)
mars.test.error <- mean((mars.pred - y2)^2)
mars.test.error
```
The regression function should be: 10.021897 + 4.853416 * h(31-bmi) + 21.54375 * h(bmi-31) * study + 2.352927 * h(height-162.4) * h(bmi-31) + 5.5933235 * h(bmi-25.2) - 5.494258 * vaccine - 2.316318 * h(weight-86.4) * h(bmi-31) + 39.271654 * h(bmi-33.8) + 14.094456 * severity * study

The test error is 251.9653

## GAM 
```{r}
set.seed(666)

gam.fit <- train(x, y, 
                 method = "gam",
                 tuneGrid = data.frame(method = "GCV.Cp", select = c(TRUE,FALSE)),
                 trControl = ctrl1)
gam.fit$bestTune
gam.fit$finalModel
```

The GAM model includes all the predictors.

Generate plot for GAM
```{r}
plot(gam.fit$finalModel)
```
We can observe that `age`, `SBP`, and `LDL` shows linear trend, whereas `bmi`, `height` and `weight` shows nonlinear trend as their df > 1.

Test error for GAM
```{r}
gam.pred <- predict(gam.fit, newdata = x2)
gam.test.error <- mean((gam.pred - y2)^2)
gam.test.error

```
The test error for GAM is 272.0012.

# Classification

Categorize `recovery_time` into two categories: < 28 days: 0,  >=28 and < 84 days: 1 since CDC states that most patients recover from acute COVID-19 illness within four weeks.
```{r}
training_set <- training_data |>
  mutate(
    recovery_time = as.factor(case_when(recovery_time < 28 ~ "normal", recovery_time >=28 ~ "long")))



```

#LDA
```{r}
set.seed(666)
ctrl2 <- trainControl(method = "repeatedcv", repeats = 10,
                     summaryFunction = twoClassSummary,
                     classProbs = TRUE)


training_set[, 1:14] <- lapply(training_set[, 1:14], as.numeric)

model.lda <- train(x = training_set[, 1:14],
                   y = training_set$recovery_time,
                   method = "lda",
                   metric = "ROC",
                   trControl = ctrl2)

testing_data[, 1:14] <- lapply(testing_data[, 1:14], as.numeric)
lda.pred2 <- predict(model.lda, newdata = testing_data, type = "prob")
head(lda.pred2)
```

# QDA
```{r}
set.seed(666)
model.qda <- train(x = training_set[, 1:14],
                   y = training_set$recovery_time,
                   method = "qda",
                   metric = "ROC",
                   trControl = ctrl2)
qda.pred2 <- predict(model.qda, newdata = testing_data, type = "prob")
head(qda.pred2)
```

Model comparison
```{r}
resamp <- resamples(list(lasso = lasso.fit, 
                         elastic_net = enet.fit,
                         pcr = pcr.fit, 
                         MARS = mars.fit,
                         GAM = gam.fit,
                         LDA = model.lda,
                         QDA = model.qda))
summary(resamp)

```

