---
title: "Midterm project"
author: "Tingyi Li"
output:
  pdf_document:
    latex_engine: xelatex
    toc: yes
    toc_depth: 2
  html_document:
    df_print: paged
    toc: yes
    toc_depth: '2'
header-includes:
- \usepackage{fancyhdr}
- \usepackage{lipsum}
- \pagestyle{fancy}
- \fancyhead[L]{HW1}
- \fancyhead[R]{Tingyi Li}
- \fancypagestyle{plain}{\pagestyle{fancy}}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, echo = T, message = FALSE, results='hide', warning=FALSE}
library(ISLR)
library(caret)
library(tidymodels)
library(pls)
library(earth)
library(ggplot2)
library(patchwork)
library(corrplot)
library(gtsummary)
```

```{r}
load("./recovery.RData")
data <- subset(dat, select = -id)
data$study <- ifelse(data$study == "A", 0, 1)

data <- data |>  
  mutate(
    gender = as.factor(gender),
    race = as.factor(race),
    smoking = as.factor(smoking),
    hypertension = as.factor(hypertension),
    diabetes = as.factor(diabetes),
    vaccine = as.factor(vaccine),
    severity = as.factor(severity),
    study = as.factor(study)
    )
```


# Data Partition
Divide data into training data (80%) and testing data (20%)
```{r}
set.seed(666)
data_split <- initial_split(data, prop = 0.8)

# training data
training_data <- training(data_split)
training_data <- na.omit(training_data)


# testing data
testing_data <- testing(data_split)
testing_data <- na.omit(testing_data)

# matrix of predictors
x <- model.matrix(recovery_time ~ ., training_data) [ ,-1]
y <- training_data$recovery_time

x2 <- model.matrix(recovery_time ~ ., testing_data) [ ,-1]
y2 <- testing_data$recovery_time

# 10-fold cross validation
ctrl1 <- trainControl(method = "cv", number = 10)
```

# EDA
## Summary Statistics
```{r}
summ_dat<-dat%>%
  select(-id)%>%
  mutate(
    gender = factor(case_when(
      gender == "1" ~ "male",  
      gender == "0" ~ "female"),
      levels = c("female", "male")
    ),
    race= factor(case_when(
      race == "1" ~ "White",  
      race == "2" ~ "Asian",
      race == "3" ~ "Black",
      race == "4" ~ "Hispanic"),
      levels = c("White", "Asian","Black","Hispanic")
    ),
    smoking=factor(case_when(
      smoking == "0" ~ "Never smoked",  
      smoking == "1" ~ "Former smoker",
      smoking == "2" ~ "Current smoker"),
      levels = c("Never smoked", "Former smoker","Current smoker")
    ),
    hypertension=factor(case_when(
      hypertension == "0" ~ "No",  
      hypertension == "1" ~ "Yes"),
      levels = c("No", "Yes")
    ),
    diabetes=factor(case_when(
      diabetes == "0" ~ "No",  
      diabetes == "1" ~ "Yes"),
      levels = c("No", "Yes")
    ),
    vaccine=factor(case_when(
      vaccine == "0" ~ "Not vaccinated",  
      vaccine == "1" ~ "Vaccinated"),
      levels = c("Not vaccinated", "Vaccinated")
    ),
    severity=factor(case_when(
      severity == "0" ~ "Not severe",  
      severity == "1" ~ "Severe"),
      levels = c("Not severe", "Severe")
    )
    )

summ_dat %>% 
  tbl_summary() %>% 
  bold_labels()%>%
  as_gt(include = everything()) %>%
  gt::tab_header("Table 1: Summary of Dataset")
```


## Feature plot for continuous variables 
```{r}
theme1 <- trellis.par.get()
theme1$plot.symbol$col <- rgb(.2, .4, .2, .5)
theme1$plot.symbol$pch <- 16
theme1$plot.line$col <- rgb(.8, .1, .1, 1)
theme1$plot.line$lwd <- 2
theme1$strip.background$col <- rgb(.0, .2, .6, .2)
trellis.par.set(theme1)

continuousData <- data|>
  dplyr::select('age', 'height', 'weight', 'bmi', 'SBP', 'LDL', 'recovery_time')
featurePlot(x = continuousData[ , 1:6],
            y = continuousData[ , 7],
            plot = "scatter",
            span = .5,
            labels = c("Predictors","Recovery Time"),
            main = "Figure 1:Relationship between Continuous Predictors and Recovery Time",
            type = c("p", "smooth"),
            layout = c(3, 2))
```

# Correlation plot
```{r}
numerical_data <- summ_dat[, sapply(summ_dat, is.numeric)]
correlation_matrix <- cor(numerical_data)

# Set up the plotting device to save the output to a file
png(filename = "./result_files/cor_plot.png", width = 8, height = 8, units = 'in', res = 300)

# Create the correlation plot
corrplot::corrplot(correlation_matrix, method = "color", addCoef.col = "black", 
                   tl.col = "black", tl.srt = 45, insig = "blank", number.cex = 0.7, diag = FALSE)

# Turn off the plotting device
dev.off()

corrplot::corrplot(correlation_matrix, method = "color", addCoef.col = "black", 
                   tl.col = "black", tl.srt = 45, insig = "blank", number.cex = 0.7, diag = FALSE)
```


## Violin plot for categorical variables
```{r}
gender_plot <- data |>
  ggplot(aes(x = gender, y = recovery_time, fill = gender))+
  geom_violin() +
  scale_x_discrete(labels = c("female", "male")) +
  labs(
       x = "Gender",
       y = "Recovery_time") +
  theme_minimal() + theme(legend.position = "none")
gender_plot

race_plot <- data |>
  ggplot(aes(x = race, y = recovery_time, fill = race)) +
  geom_violin() +
  scale_x_discrete(labels = c("White", "Asian", "Black", "Hispanic")) +
  labs(
       x = "race",
       y = "Recovery_time") +
  theme_minimal() + theme(legend.position = "none")
race_plot

smoking_plot <- data |>
  ggplot(aes(x = smoking, y = recovery_time, fill = smoking)) +
  geom_violin() +
  scale_x_discrete(labels = c("Never", "Former", "Current")) +
  labs(
       x = "smoking",
       y = "Recovery_time") +
  theme_minimal() + theme(legend.position = "none")
smoking_plot 

hypertension_plot <- data |>
  ggplot(aes(x = hypertension, y = recovery_time, fill = hypertension)) +
  geom_violin() +
  scale_x_discrete(labels = c("No", "Yes")) +
  labs(
       x = "hypertension",
       y = "Recovery_time") +
  theme_minimal() + theme(legend.position = "none")
hypertension_plot

diabetes_plot <- data |>
  ggplot(aes(x = diabetes, y = recovery_time, fill = diabetes)) +
  geom_violin() +
  scale_x_discrete(labels = c("No", "Yes")) +
  labs(
       x = "diabetes",
       y = "Recovery_time") +
  theme_minimal() + theme(legend.position = "none")
diabetes_plot

vaccine_plot <- data |>
   ggplot(aes(x = vaccine, y = recovery_time, fill = vaccine)) +
   geom_violin() +
   scale_x_discrete(labels = c("Not vaccinated", "Vaccinated")) +
   labs(
        x = "vaccine",
        y = "Recovery_time") +
   theme_minimal() + theme(legend.position = "none")
vaccine_plot

severity_plot <- data |>
  ggplot(aes(x = severity, y = recovery_time, fill = severity)) +
  geom_violin() +
  scale_x_discrete(labels = c("Not severe", "Severe")) +
  labs(
       x = "severity",
       y = "Recovery_time") +
  theme_minimal() + theme(legend.position = "none")
severity_plot

study_plot <- data |>
  ggplot(aes(x = study, y = recovery_time, fill = study)) +
  geom_violin() +
  scale_x_discrete(labels = c("A", "B")) +
  labs(
       x = "study",
       y = "Recovery_time") +
  theme_minimal() + theme(legend.position = "none")
study_plot

combined <- gender_plot + race_plot + smoking_plot + hypertension_plot + diabetes_plot + vaccine_plot + severity_plot + study_plot + plot_layout(ncol = 3)

combined + plot_annotation(title = "Figure 2: Relationship between Categorical Predictors and Recovery Time")
```




# Linear models

## Fit a lasso model
```{r}
set.seed(666)

lasso.fit <- train(recovery_time ~ .,
                   data = training_data,
                   method = "glmnet",
                   tuneGrid = expand.grid(alpha = 1, 
                                          lambda = exp(seq(-25, 5, length = 100))),
                   trControl = ctrl1)
plot(lasso.fit, xTrans = log)

## Select the best tuning parameter
lasso.fit$bestTune

# coefficients in the final model
coef(lasso.fit$finalModel, lasso.fit$bestTune$lambda)
```
The selected tuning parameter is 0.00912288.

Test errors
```{r}
set.seed(666)
lasso.pred <- predict(lasso.fit, newdata = testing_data)
mean((lasso.pred - testing_data[, "recovery_time"])^2)
```

The test error is 298.3016.


## Fit an elastic net model
```{r}
set.seed(666)
enet.fit <- train(recovery_time ~ .,
                  data = training_data,
                  method = "glmnet",
                  tuneGrid = expand.grid(alpha = seq(0, 1, length = 21), 
                                         lambda = exp(seq(-25, 5, length = 100))),

                  trControl = ctrl1)
## Select the best tuning parameter
enet.fit$bestTune

#Plot
myCol <- rainbow(25)
myPar <- list(superpose.symbol = list(col = myCol),
              superpose.line = list(col = myCol))

plot(enet.fit, par.settings = myPar)

# coefficients in the final model
coef(enet.fit$finalModel, enet.fit$bestTune$lambda)

# test error
enet.pred <- predict(enet.fit, newdata = testing_data)
mean((enet.pred - testing_data[, "recovery_time"])^2)
```
The selected tuning parameter is 0.00367552, and the test error is 297.1642.


## Fit a PCR model
```{r}
set.seed(666)
pcr.fit <- train(x, y,
                 method = "pcr",
                 tuneGrid = data.frame(ncomp = 1:15),
                 trControl = ctrl1,
                 scale = TRUE)
predy2.pcr2 <- predict(pcr.fit, newdata = x2)
mean((y2 - predy2.pcr2)^2)

ggplot(pcr.fit, highlight = TRUE)
summary(pcr.fit)
```
There are 15 components in the model and the test error is 323.779.

# Nonlinear models
## Multivariate Adaptive Regression Spline (MARS)
```{r}
set.seed(666)
mars.grid <- expand.grid(degree = 1:4,
                         nprune = 2:38)
mars.fit <- train(x, y,
                  method = "earth",
                  tuneGrid = mars.grid,
                  trControl = ctrl1)

plot(mars.fit)

# best tune
mars.fit$bestTune

coef(mars.fit$finalModel)
```

Test error for MARS
```{r}
mars.pred <- predict(mars.fit, newdata = x2)
mars.test.error <- mean((mars.pred - y2)^2)
mars.test.error
```
The regression function should be: -3.1983530 + 6.3999877 * h(31-bmi) + 25.6820131 * h(bmi-31) * study1 + 7.9260754 * h(bmi-25.2) - 0.6277843 * h(weight-86.4) * h(bmi-31) 

The test error is 279.0367

## GAM 
```{r}
set.seed(666)

gam.fit <- train(x, y, 
                 method = "gam",
                 tuneGrid = data.frame(method = "GCV.Cp", select = c(TRUE,FALSE)),
                 trControl = ctrl1)
gam.fit$bestTune
gam.fit$finalModel
```

The GAM model includes all the predictors.

Generate plot for GAM
```{r}
plot(gam.fit$finalModel)
```


Test error for GAM
```{r}
gam.pred <- predict(gam.fit, newdata = x2)
gam.test.error <- mean((gam.pred - y2)^2)
gam.test.error

```
The test error for GAM is 272.0012.



# Model comparison
```{r}
resamp <- resamples(list(lasso = lasso.fit, 
                         elastic_net = enet.fit,
                         pcr = pcr.fit, 
                         MARS = mars.fit,
                         GAM = gam.fit))
summary(resamp)

bwplot(resamp, metric = "RMSE")
```

Mars model should be selected as the final model since it has the lowest RMSE value.
