---
title: "Yvonne_Chen_sc5078_midterm_project"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, echo = T, message = FALSE, results='hide', warning=FALSE}
library(caret)
library(tidymodels)
library(splines)
library(mgcv)
library(pdp)
library(earth)
library(tidyverse)
library(ggplot2)
library(dplyr)
library(discrim) 
library(MASS)
library(mlbench)
library(pROC)
library(klaR)
library(plotmo)
library(patchwork)
library(corrplot)
library(gtsummary)
```

# Dataset split
Randomly split dataset into traning (80%) and testing data (20%).
```{r}
# load dataset
load("~/Desktop/8106 Data Science II/data science II midterm project/recovery.RData")

# remove id
data <- subset(dat, select = -id)

# set study A/B to 0/1
data$study <- ifelse(data$study == "A", 0, 1)


set.seed(666)
data_split <- initial_split(data, prop = 0.8)

# training data
training_data <- training(data_split)

# test data
testing_data <- testing(data_split)

# matrix of predictors
x <- model.matrix(recovery_time ~ ., training_data) [ ,-1]
x2 <- model.matrix(recovery_time ~ ., testing_data) [ ,-1]
y <- training_data$recovery_time

```

# EDA
## Summary Statistics
```{r}
summ_dat<-dat%>%
  select(-id)%>%
  mutate(
    gender = factor(case_when(
      gender == "1" ~ "male",  
      gender == "0" ~ "female"),
      levels = c("female", "male")
    ),
    race= factor(case_when(
      race == "1" ~ "White",  
      race == "2" ~ "Asian",
      race == "3" ~ "Black",
      race == "4" ~ "Hispanic"),
      levels = c("White", "Asian","Black","Hispanic")
    ),
    smoking=factor(case_when(
      smoking == "0" ~ "Never smoked",  
      smoking == "1" ~ "Former smoker",
      smoking == "2" ~ "Current smoker"),
      levels = c("Never smoked", "Former smoker","Current smoker")
    ),
    hypertension=factor(case_when(
      hypertension == "0" ~ "No",  
      hypertension == "1" ~ "Yes"),
      levels = c("No", "Yes")
    ),
    diabetes=factor(case_when(
      diabetes == "0" ~ "No",  
      diabetes == "1" ~ "Yes"),
      levels = c("No", "Yes")
    ),
    vaccine=factor(case_when(
      vaccine == "0" ~ "Not vaccinated",  
      vaccine == "1" ~ "Vaccinated"),
      levels = c("Not vaccinated", "Vaccinated")
    ),
    severity=factor(case_when(
      severity == "0" ~ "Not severe",  
      severity == "1" ~ "Severe"),
      levels = c("Not severe", "Severe")
    )
    )

summ_dat %>% 
  tbl_summary() %>% 
  bold_labels()%>%
  as_gt(include = everything()) %>%
  gt::tab_header("Table 1: Summary of Dataset")
```


## Feature plot for continuous variables 
```{r}
theme1 <- trellis.par.get()
theme1$plot.symbol$col <- rgb(.2, .4, .2, .5)
theme1$plot.symbol$pch <- 16
theme1$plot.line$col <- rgb(.8, .1, .1, 1)
theme1$plot.line$lwd <- 2
theme1$strip.background$col <- rgb(.0, .2, .6, .2)
trellis.par.set(theme1)

continuousData <- data|>
  dplyr::select('age', 'height', 'weight', 'bmi', 'SBP', 'LDL', 'recovery_time')
featurePlot(x = continuousData[ , 1:6],
            y = continuousData[ , 7],
            plot = "scatter",
            span = .5,
            labels = c("Predictors","Recovery Time"),
            main = "Figure 1:Relationship between Continuous Predictors and Recovery Time",
            type = c("p", "smooth"),
            layout = c(3, 2))
```

# Correlation plot
```{r}
numerical_data <- summ_dat[, sapply(summ_dat, is.numeric)]
correlation_matrix <- cor(numerical_data)

# Create the correlation plot
corrplot::corrplot(correlation_matrix, method = "color", addCoef.col = "black", 
                   tl.col = "black", tl.srt = 45, insig = "blank", number.cex = 0.7, diag = FALSE)


corrplot::corrplot(correlation_matrix, method = "color", addCoef.col = "black", 
                   tl.col = "black", tl.srt = 45, insig = "blank", number.cex = 0.7, diag = FALSE)
```


## Violin plot for categorical variables
```{r}
gender_plot <- data |>
  ggplot(aes(x = gender, y = recovery_time, fill = gender))+
  geom_violin() +
  scale_x_discrete(labels = c("female", "male")) +
  labs(
       x = "Gender",
       y = "Recovery_time") +
  theme_minimal() + theme(legend.position = "none")
gender_plot

race_plot <- data |>
  ggplot(aes(x = race, y = recovery_time, fill = race)) +
  geom_violin() +
  scale_x_discrete(labels = c("White", "Asian", "Black", "Hispanic")) +
  labs(
       x = "race",
       y = "Recovery_time") +
  theme_minimal() + theme(legend.position = "none")
race_plot

smoking_plot <- data |>
  ggplot(aes(x = smoking, y = recovery_time, fill = smoking)) +
  geom_violin() +
  scale_x_discrete(labels = c("Never", "Former", "Current")) +
  labs(
       x = "smoking",
       y = "Recovery_time") +
  theme_minimal() + theme(legend.position = "none")
smoking_plot 

hypertension_plot <- data |>
  ggplot(aes(x = hypertension, y = recovery_time, fill = hypertension)) +
  geom_violin() +
  scale_x_discrete(labels = c("No", "Yes")) +
  labs(
       x = "hypertension",
       y = "Recovery_time") +
  theme_minimal() + theme(legend.position = "none")
hypertension_plot

diabetes_plot <- data |>
  ggplot(aes(x = diabetes, y = recovery_time, fill = diabetes)) +
  geom_violin() +
  scale_x_discrete(labels = c("No", "Yes")) +
  labs(
       x = "diabetes",
       y = "Recovery_time") +
  theme_minimal() + theme(legend.position = "none")
diabetes_plot

vaccine_plot <- data |>
   ggplot(aes(x = vaccine, y = recovery_time, fill = vaccine)) +
   geom_violin() +
   scale_x_discrete(labels = c("Not vaccinated", "Vaccinated")) +
   labs(
        x = "vaccine",
        y = "Recovery_time") +
   theme_minimal() + theme(legend.position = "none")
vaccine_plot

severity_plot <- data |>
  ggplot(aes(x = severity, y = recovery_time, fill = severity)) +
  geom_violin() +
  scale_x_discrete(labels = c("Not severe", "Severe")) +
  labs(
       x = "severity",
       y = "Recovery_time") +
  theme_minimal() + theme(legend.position = "none")
severity_plot

study_plot <- data |>
  ggplot(aes(x = study, y = recovery_time, fill = study)) +
  geom_violin() +
  scale_x_discrete(labels = c("A", "B")) +
  labs(
       x = "study",
       y = "Recovery_time") +
  theme_minimal() + theme(legend.position = "none")
study_plot

combined <- gender_plot + race_plot + smoking_plot + hypertension_plot + diabetes_plot + vaccine_plot + severity_plot + study_plot + plot_layout(ncol = 3)

combined + plot_annotation(title = "Figure 2: Relationship between Categorical Predictors and Recovery Time")
```

# Nonlinear Methods
## Multivariate Adaptive Regression Splines (MARS)
Train a multivariate adaptive regression spline (MARS) model.
```{r}
mars_grid <- expand.grid(degree = 1:4, 
                         nprune = 2:38)

ctrl1 <- trainControl(method = "cv", number = 10)

set.seed(666)
mars.fit <- train(x, y,
                  method = "earth",
                  tuneGrid = mars_grid,
                  trControl = ctrl1)

ggplot(mars.fit)

mars.fit$bestTune

coef(mars.fit$finalModel) 

```
The test error is 279.0367.
```{r}
# prediction
mars.pred <- predict(mars.fit, newdata = testing_data)
#test error
mean((mars.pred - testing_data[, "recovery_time"])^2)
```

## Generalized additive model (GAM)
```{r}
set.seed(666)
gam.fit <- train(x, y,
                 method = "gam",
                 tuneGrid = data.frame(method = "GCV.Cp", select = c(TRUE,FALSE)),
                 trControl = ctrl1)

gam.fit$bestTune

gam.fit$finalModel
```
The GAM model include all the predictors.

```{r}
plot(gam.fit$finalModel)
```
According to plots, variables `age`, `SBP`, `LDL`, and `weight` are more likely to have linear relationship while variables `bmi` and `height` are more likely to have nonlinear relationship.

The test error is 272.0012.
```{r}
# prediction
gam.pred <- predict(gam.fit, newdata = x2)
#test error
mean((gam.pred - testing_data[, "recovery_time"])^2)
```

# Linear Methods
## lasso model
```{r}
set.seed(666)
lasso.fit <- train(recovery_time ~ .,
                   data = training_data,
                   method = "glmnet",
                   tuneGrid = expand.grid(alpha = 1,
                                          lambda = exp(seq(5, -25, length = 100))),
                   trControl = ctrl1)

plot(lasso.fit, xTrans = log)

lasso.fit$bestTune

# coefficients in the final model w/ min CV MSE
coef_min <- coef(lasso.fit$finalModel, s = lasso.fit$bestTune$lambda)

sum(coef_min[-1, ] != 0)
```
There are 17 variables in the final lasso model. The selected tunning parameter applying minimum CV MSE: lambda = 0.00912288.

The test error is 298.3016
``` {r}
# prediction
lasso.pred <- predict(lasso.fit, newdata = testing_data)

# test error
mean((lasso.pred - testing_data[, "recovery_time"])^2)
```

## elastic net model
```{r}
set.seed(666)
# min CV MSE
enet.fit <- train(recovery_time ~ .,
                data = training_data,
                method = "glmnet",
                tuneGrid = expand.grid(alpha = seq(0, 1, length = 21),
                                       lambda = exp(seq(-25, 5, length = 100))),
                trControl = ctrl1)
enet.fit$bestTune
coef(enet.fit$finalModel, enet.fit$bestTune$lambda)
enet.pred <- predict(enet.fit, newdata = testing_data)

# plot
myCol <- rainbow(25)
myPar <- list(superpose.symbol = list(col = myCol),
              superpose.line = list(col = myCol))

plot(enet.fit, par.settings = myPar)

# test error
mean((enet.pred - testing_data[, "recovery_time"])^2)
```
alpha = 0.15, lambda = 0.00367552, test error = 297.1642

## principal components regression (PCR)
```{r}
set.seed(666)
pcr.fit <- train(x, y,
                 method = "pcr",
                 tuneGrid = data.frame(ncomp = 1:15),
                 trControl = ctrl1,
                 preProcess = c("center", "scale"))

x2 <- model.matrix(recovery_time ~ ., testing_data) [, -1]
predy.pcr <- predict(pcr.fit, newdata = x2)

# test error
mean((testing_data$recovery_time - predy.pcr)^2)

ggplot(pcr.fit,highlight = TRUE)

# ncomp
pcr.fit$finalModel$ncomp


```
The test error is 323.779
There are 15 components included in the model.

# Classification
According to CDC, most patients appear to recover from acute COVID-19 illness within 4 weeks, which is 28 days. Classify outcome `recovery_time` into < 28 days: normal, >= 28 days: long.
```{r}
training_class <- training_data %>%
  mutate(recovery_time = as.factor(ifelse(recovery_time < 28, "normal", "long")))
```
## LDA
```{r}
# Exploratory analysis: LDA based on every combination of two variables
partimat(recovery_time ~ age + height + weight + bmi + SBP + LDL, 
         data = training_class, method = "lda")
```

```{r}
set.seed(666)
ctrl2 <- trainControl(method = "repeatedcv", repeats = 10,
                     summaryFunction = twoClassSummary,
                     classProbs = TRUE)

# Convert non-numeric columns to numeric
training_class[, 1:14] <- lapply(training_class[, 1:14], as.numeric)

model.lda <- train(x = training_class[, 1:14],
                   y = training_class$recovery_time,
                   method = "lda",
                   metric = "ROC",
                   trControl = ctrl2)

testing_data2 <- testing_data
testing_data2[, 1:14] <- lapply(testing_data2[, 1:14], as.numeric)

lda.pred <- predict(model.lda, newdata = testing_data2, type = "prob")
head(lda.pred)
```

# QDA
```{r}
model.qda <- train(x = training_class[, c("age", "height", "weight", "bmi", "SBP", "LDL")],
                   y = training_class$recovery_time,
                   method = "lda",
                   metric = "ROC",
                   trControl = ctrl2)

qda.pred <- predict(model.qda, newdata = testing_data, type = "prob")
head(lda.pred)
```

# Model Comparison
```{r}
resamp <- resamples(list(mars = mars.fit,
                      gam = gam.fit,
                      enet = enet.fit,
                      lasso = lasso.fit,
                      pcr = pcr.fit))

summary(resamp)

bwplot(resamp, metric = "RMSE")
```

